---
title: "Add indications to Drug Utilisation Cohort"
author: "Marti Catala, Mike Du, Yuchen Guo, Kim Lopez-Guell, Edward Burn, Xintong Li"
date: "`r Sys.Date()`"
output: 
  html_document:
    pandoc_args: [
      "--number-offset=1,0"
      ]
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Add indications to Drug Utilisation Cohort}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
if (Sys.getenv("EUNOMIA_DATA_FOLDER") == "") Sys.setenv("EUNOMIA_DATA_FOLDER" = tempdir())
if (!dir.exists(Sys.getenv("EUNOMIA_DATA_FOLDER"))) dir.create(Sys.getenv("EUNOMIA_DATA_FOLDER"))
if (!CDMConnector::eunomia_is_available()) CDMConnector::downloadEunomiaData()
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

In this vignette, we demonstrate how to use the `addIndications()` function to establish a binary indicator between the drug utilisation cohort and another concept-based cohort. 

The DrugUtilisation package is designed to work with data in the OMOP CDM format, so our first step is to create a reference to the data using the DBI and CDMConnector packages. The connection to a Postgres database would look like:

```{r setup , message= FALSE, warning=FALSE}
library(DrugUtilisation)
library(DBI)
library(duckdb)
library(CDMConnector)
library(CodelistGenerator)
library(dplyr)
library(PatientProfiles)


con <- DBI::dbConnect(duckdb::duckdb(), eunomia_dir())

cdm <- CDMConnector::cdm_from_con(
  con = con,
  cdm_schema = "main",
  write_schema = "main"
)
```

## Create a drug utilisation cohort

We will use *Acetaminophen* as our example drug to construct our drug utilisation cohort. To begin, we'll employ the CodelistGenerator package to generate a concept list associated with *Acetaminophen*.

```{r, eval=TRUE, message= FALSE, warning=FALSE}
conceptList <- CodelistGenerator::getDrugIngredientCodes(cdm, "acetaminophen")
conceptList
```

Next, we can create a drug utilisation cohort by using the `conceptList` with the `generateDrugUtilisationCohortSet()` function. For a better understanding of the arguments and functionalities of `generateDrugUtilisationCohortSet()`, please refer to the *Use DrugUtilisation to create a cohort* vignette.

```{r, eval=TRUE, message= FALSE, warning=FALSE}
cdm <- generateDrugUtilisationCohortSet(
  cdm = cdm,
  name = "acetaminophen_users",
  conceptSet = conceptList,
  limit = "All",
  gapEra = 30,
  priorUseWashout = 0
)
```
## Create a indication cohort

Next we going to create our indications cohort to indicate patients with sinusitis and bronchitis. This can be done by using `CDMConnector::generateConceptCohortSet()`.

```{r, eval=TRUE, message= FALSE, warning=FALSE}
indications <-
  list(
    sinusitis = c(257012, 4294548, 40481087),
    bronchitis = c(260139, 258780)
  )

cdm <-
  CDMConnector::generateConceptCohortSet(cdm, name = "indications_cohort", indications)

cohortCount(cdm[["indications_cohort"]]) %>%
  left_join(
    settings(cdm[["indications_cohort"]]) %>%
      select(cohort_definition_id, cohort_name),
    by = "cohort_definition_id"
  )

```

# Add indications with addIndication() function

Then to add indication to the drug utilisation cohort we can simple use the `addIndication()` function. To do that, we only need the drug utilisation cohort and the `indicationCohortName`. However, other arguments can be specified using the additional parameters. An example is provided below. 

```{r, eval=TRUE, message= FALSE, warning=FALSE}

cdm[["acetaminophen_users"]] |>
  addIndication(
    indicationCohortName = "indications_cohort",
    indicationWindow =  list(c(0,0),c(-30,0),c(-365,0)),
    unknownIndicationTable =  c("condition_occurrence")
  )

```

Use `indicationGap` to specify the indication gaps, which are defined as the gap between the event and the indication. Additionally, you can use `unknownIndicationTable` to specify the tables to look for unknown indication. 

## Summarise indications with summariseIndication()
To create a summary table of the indications cohort, you can use the `summariseIndication()` function.

```{r, eval=TRUE, message= FALSE, warning=FALSE}
# cdm[["acetaminophen_users"]] %>%
#   addIndication(
#     cdm = cdm, 
#     indicationCohortName = "indications_cohort",
#     indicationGap =  c(0, 30, 365),
#     unknownIndicationTable =  c("condition_occurrence")
#   ) %>%
#   summariseIndication(cdm) %>%
#   select("variable_name", "estimate_name", "estimate_value")

```

You can also summarise the indications by using the `strata` argument in the `summariseIndication()` function. In the example below, it is summarized by `ageGroup` and `sex`.

```{r, eval=TRUE, message= FALSE, warning=FALSE}

# cdm[["acetaminophen_users"]] %>%
#   addDemographics(ageGroup = list(c(0, 19), c(20, 150))) %>%
#   addIndication(
#     cdm = cdm,
#     indicationCohortName = "indications_cohort",
#     indicationGap =  c(0),
#     unknownIndicationTable =  c("condition_occurrence")
#   ) %>%
#   summariseIndication(
#     cdm,
#     strata = list("age" = "age_group", "sex" = "sex")) %>%
#       select("variable_name", "estimate_name", "estimate_value","strata_name")

```
