---
title: "Use DrugUtilisation to create a cohort"
author: "Martí Català, Mike Du, Yuchen Guo, Kim Lopez-Guell, Edward Burn, Xintong Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use DrugUtilisation to create a cohort}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r , include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

In this vignette we will introduce how to create a drug users cohorts. A cohort is a set of people that satisfy a certain inclusion criteria durng a certain time frame. The cohort object is defined in : `vignette("a03_cohort_table", package = "omopgenerics")`.

The function `generateDrugUtilisationCohortSet` is used to generate cohorts of drug users based on the drug_exposure table and a conceptSet.

These cohorts can be subsetted to the exposures of interest applying the different inclusion criteria:

- Require that entries are in a certain date range `requireDrugInDateRange()`.

- Subset to the first entry `requireIsFirstDrugEntry()`.

- Require a certain time in observation before the entries `requireObservationBeforeDrug()`.

- Require a certain time before exposure `requirePriorDrugWashout()`.

# Creating a `cdm_reference` object

The first thing that we need is a cdm reference object to our OMOP CDM instance. You can learn how to create cdm references using CDMConnector here: CDMConnector::a04_DBI_connection_examples

The DrugUtilisation packages contains some mock data that can be useful to test the package:

```{r}
library(DrugUtilisation)

cdm <- mockDrugUtilisation(numberIndividuals = 100, seed = 1)

cdm
```

# Create a drug users cohort

To create a basic drug users cohort we need two things:

- A conceptSet: will determine which concepts we will use.
- A gapEra: will determine how we will collapse those exposures.

## Creating a conceptSet

There are four possible outputs for a conceptSet:

- A named list of concept ids
```{r}
conceptSet <- list(acetaminophen = c(1, 2, 3))
conceptSet
```

- A `codelist` object, see omopgenerics "a02_concept_set"
```{r}
conceptSet <- list(acetaminophen = c(1, 2, 3)) |> omopgenerics::newCodelist()
conceptSet
conceptSet$acetaminophen
```

- A `codelist_with_details` object, see omopgenerics "a02_concept_set"
```{r}
conceptSet <- list(acetaminophen = dplyr::tibble(
  concept_id = c(1125315, 1125360),
  concept_name = c("acetaminophen", "acetaminophen 500 MG Oral Capsule"),
  domain_id = "Drug",
  vocabulary_id = "RxNorm",
  standard_concept = "S"
)) |> 
  omopgenerics::newCodelistWithDetails()
conceptSet
conceptSet$acetaminophen
```

- A `conceptSetExpression` object, see omopgenerics "a02_concept_set"

```{r}
conceptSet <- list(acetaminophen = dplyr::tibble(
  concept_id = 1125315, 
  excluded = FALSE,
  descendants = TRUE,
  mapped = FALSE
)) |> 
  omopgenerics::newConceptSetExpression()
conceptSet
conceptSet$acetaminophen
```

The package [CodelistGenerator](https://cran.r-project.org/package=CodelistGenerator)
can be very useful to create conceptSet.

```{r}
library(CodelistGenerator)
```

For example we can create a conceptSet based in an ingredient with `getDrugIngredientCodes()`:
```{r}
codes <- getDrugIngredientCodes(cdm = cdm, name = "acetaminophen")
codes[["acetaminophen"]]
```

We could also use the function `codesFromConceptSet()` to read a concept set from a json file:
```{r}
codes <- codesFromConceptSet(path = system.file("acetaminophen.json", package = "DrugUtilisation"), cdm = cdm)
codes
```

## The gapEra parameter

```{r}

```


## Create your cohort

We will then create now a cohort with all the drug users of acetaminophen with a gapEra of 30 days.

```{r, messages = TRUE}
codes <- getDrugIngredientCodes(cdm = cdm, name = "acetaminophen")
cdm <- generateDrugUtilisationCohortSet(cdm = cdm, "acetaminophen_cohort", conceptSet = codes, gapEra = 30)
cdm
```

We can compare what we see with what we would expect if we look at the individual with more records we can see how all of them are joined into a single exposure as the records overlap each other:

```{r}
cdm$drug_exposure |>
  dplyr::filter(drug_concept_id %in% codes$acetaminophen & person_id == 69)
```

```{r}
cdm$acetaminophen_cohort |>
  dplyr::filter(subject_id == 69)
```

In this case gapEra did not have a big impact as we can see in the attrition:

```{r}
attrition(cdm$acetaminophen_cohort)
```

We can see this particular case of this individual:

```{r}
cdm$drug_exposure |>
  dplyr::filter(drug_concept_id %in% codes$acetaminophen & person_id == 50)
```

In this case we have 3 exposures separated by 3 days, so if we use the 30 days gap both exposures are joined into a single episode, wheras if we would use a gapEra smaller than 3 we would consider them as different episodes.

```{r}
cdm$acetaminophen_cohort |>
  dplyr::filter(subject_id == 50)
```

We can access the other cohort attributes using the adequate functions. In settings we can see that the gapEra used is recorded or with cohortCodelist we can see which was the codelisy used to create the cohort.

```{r}
settings(cdm$acetaminophen_cohort)
cohortCount(cdm$acetaminophen_cohort)
cohortCodelist(cdm$acetaminophen_cohort, cohortId = 1)
```

# Apply inclusion criteria to drug cohorts

# The order matters

# Alternatives to create cohorts

`generateDrugUtilisationCohortSet()` is a very specif function to create drug users cohorts. Most of the functions in this package need a cohort as input, this cohort will usually be a cohort generated using `generateDrugUtilisationCohortSet()`, but there exist some alternative to create cohorts

- You can create your own custom cohorts with `tidyverse` functions and `omopgenerics::newCohortTable()`. See omopgenerics vignette.

- You can create concept based cohorts using the `OHDSI` package `CohortConstructor`. See CohortConstructor "a01_building_base_cohorts".

- You can create cohorts based on ATLAS definitions using CDMConnector. "a02_cohorts"
