---
title: "Characterise a cohort using Drug Utilisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Characterise a cohort using Drug Utilisation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
if (Sys.getenv("EUNOMIA_DATA_FOLDER") == "") Sys.setenv("EUNOMIA_DATA_FOLDER" = tempdir())
if (!dir.exists(Sys.getenv("EUNOMIA_DATA_FOLDER"))) dir.create(Sys.getenv("EUNOMIA_DATA_FOLDER"))
if (!CDMConnector::eunomia_is_available()) CDMConnector::downloadEunomiaData()
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

In DrugUtilisation package, there are also functions on daily dose of
drugs. The best way to learn these functions is to consider a few
examples.

## Create mock data first

```{r, message= FALSE, warning=FALSE}
library(DrugUtilisation)
library(CodelistGenerator)
library(CDMConnector)
library(dplyr)
library(PatientProfiles)

cdm <- mockDrugUtilisation(numberIndividual  = 200)
```

## Add daily dose information via addDailyDose() function

The following example calculates the daily dose of *acetaminophen* in
the `drug_exposure` table.

```{r, message= FALSE, warning=FALSE}
cdm$drug_exposure |>
  filter(drug_concept_id == 2905077) |>
  addDailyDose(ingredientConceptId = 1125315) |>
  glimpse()
```

The `drug_exposure` table is first subsetted according to whether
`drug_concept_id` in the `drug_exposure` table equals $2905077$ or not;
this is to subset the `drug_exposure` table to the use of *acetaminophen*
in the marketed product form. Later the ID $1125315$ is supplied in
`addDailyDose()` function, which corresponds to the ingredient form of
*acetaminophen*.

We see that there are two extra columns at the end detailing the daily
dose of *acetaminophen* - `daily_dose` and `unit`. Take first entry for
example, we see that in that drug exposure of *acetaminophen*, the daily
dose on average is $48,000$ milligram. The way daily dose is being
calculated is to take into consideration both the strength of the drug
and the duration of the episode.

## Add route information via addRoute() function

Next, the user could also obtain the information on the route of each
drug used in the `drug_exposure` table.

```{r, message= FALSE, warning=FALSE}
cdm$drug_exposure |>
  addRoute() |>
  glimpse()
```

Similarly, this adds one extra column named `route`. For example, the
drug used in the first entry of the `drug_exposure` table is an oral
drug.

## Finding out the pattern information using patternTable() function

The user could also find the patterns used in the `drug_strength` table.
The output will also include a column of potentially valid and invalid
combinations. The idea of a pattern to provide a platform to associate
each drug in the `drug_strength` table with its constituent ingredients.

```{r, message= FALSE, warning=FALSE}
patternTable(cdm) |>
  glimpse()
```

The output has three important columns, namely `number_concepts`,
`number_ingredients` and `number_records`, which corresponds to count of
distinct concepts in the patterns, count of distinct ingredients
involved and overall count of records in the patterns respectively.

We also see that there is a column named `pattern_id` together with
information such as `number_concepts` and `number_ingredients`. The idea
is to use this output in conjunction with the data named
`patternWithFormula` that the user can find in this package via
`pattern_id`. Please see the data `patternWithFormula` for different
patterns, their associated formula and combinations of `amount_unit`,
`numerator_unit` and `denominator_unit`.

```{r, message= FALSE, warning=FALSE}
patternsWithFormula |>
  glimpse()
```

## Finding out the dose coverage using summariseDoseCoverage() function

This package also provides a function to check the coverage of daily
dose computation for chosen concept sets and ingredients. Again let's
take *acetaminophen* as an example.

```{r, message= FALSE, warning=FALSE}
summariseDoseCoverage(cdm = cdm, ingredientConceptId = 1125315) |>
  glimpse()
```

The output will summarise the usage of *acetaminophen* in the database.
For example, overall there are $13$ records of *acetaminophen* use and
none missing in terms of the duration and strength. By default the
output will also include the mean, median, lower and upper quartiles and
standard deviation of the daily dose of *acetaminophen* (calculated using
`addDailyDose()`). The results will also be stratified by unit, route
(which we saw in `addRoute()` function) and pattern (which we saw in
`patternTable()` function).

One may wish to display the output of `summariseDoseCoverage()` in a gt
table form like so:

```{r, message= FALSE, warning=FALSE}
summariseDoseCoverage(cdm = cdm, ingredientConceptId = 1125315) |>
  tableDoseCoverage()
```

The user also has the freedom to customize the gt table output. For
example the following will suppress the `cdmName`:

```{r, message= FALSE, warning=FALSE}
summariseDoseCoverage(cdm = cdm, ingredientConceptId = 1125315) |>
  tableDoseCoverage(cdmName = F)
```

Additionally, if the user wants to specify a title for the gt output,
they could try the following:

```{r, message= FALSE, warning=FALSE}
summariseDoseCoverage(cdm = cdm, ingredientConceptId = 1125315) |>
  tableDoseCoverage(.options = list(title = "Title of summariseDoseCoverage"))
```
