---
title: "Daily dose calculation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Daily dose calculation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
if (Sys.getenv("EUNOMIA_DATA_FOLDER") == "") Sys.setenv("EUNOMIA_DATA_FOLDER" = tempdir())
if (!dir.exists(Sys.getenv("EUNOMIA_DATA_FOLDER"))) dir.create(Sys.getenv("EUNOMIA_DATA_FOLDER"))
if (!CDMConnector::eunomia_is_available()) CDMConnector::downloadEunomiaData()
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

In this vignette is assessed how daily dose is calculated in the DrugUtilisation package. This function is used internally in `addDrugUtilisation()`.

## Daily dose

Daily dose is always computed at the ingredient level. So we can calculate the daily dose for each record in *drug exposure* table for each given ingredient. Then the first step to calculate the daily dose for a given drug record and an ingredient concept id is to examine the relationship between drug concept id and ingredient concept id through the *drug strength* table:

```{r}
library(DrugUtilisation)
cdm <- mockDrugUtilisation(numberIndividuals = 100, seed = 123456)
cdm$drug_strength |>
  dplyr::glimpse()
```

You can read the documentation of the *drug strength* table and description of the different fields here: <https://www.ohdsi.org/web/wiki/doku.php?id=documentation:cdm:drug_strength>.

Not all drug concept ids and ingredient concept ids can be related, if no relation is found then daily dose is considered as `NA`.

Using vocabulary version: "v5.0 31-AUG-23" there exist 2,980,115 relationships between a drug concept id and an ingredient concept id. These relationships can be classified into **128** different patterns. Patterns are identified of combinations of 6 variables:

- *amount*: Whether the amount_value field is numeric or NA.
- *amount_unit*: The unit of the amount field.
- *numerator*: Whether the numerator_value field is numeric or NA.
- *numerator_unit*: The unit of the numerator field.
- *denominator*: Whether the denominator_value field is numeric or NA.
- *denominator_unit*: The unit of the denominator field.

These 128 combinations were analysed to see if they could be used to compute daily dose. **41 viable** patterns were identified, these patterns covered a total of 2,514,608 (84%) relationships between drug concept id and ingredient concept id. The patterns were classified into 4 different formulas:

1. **Time based with denominator**

This formula was applied for the following 3 patterns that cover 8,044 (<1%) relationships:

```{r, echo = FALSE}
patternsWithFormula |>
  dplyr::filter(formula_name == "time based with denominator") |>
  dplyr::select(pattern_id, amount, amount_unit, numerator, numerator_unit, denominator, denominator_unit) |>
  flextable::flextable()
```

The formula in this case will be as follows:

\begin{equation}
if (denominator > 24) \rightarrow \textrm{daily dose} = 24 \cdot \frac{numerator}{denominator} \\
if (denominator \leq 24) \rightarrow \textrm{daily dose} = numerator
\end{equation}

Note that daily dose has always unit associated in this case it will be determined by the `numerator_unit` field.

2. **Time based no denominator**

This formula was applied for the following 2 patterns that cover 5,611 (<1%) relationships:

```{r, echo = FALSE}
patternsWithFormula |>
  dplyr::filter(formula_name == "time based no denominator") |>
  dplyr::select(pattern_id, amount, amount_unit, numerator, numerator_unit, denominator, denominator_unit) |>
  flextable::flextable()
```

The formula in this case will be as follows:

\begin{equation}
\textrm{daily dose} = 24 \cdot numerator
\end{equation}

In this case unit will be determined by the `numerator_unit` field.

3. **Fixed amount formulation**

This formula was applied for the following 6 patterns that cover 1,102,435 (37%) relationships:

```{r, echo = FALSE}
patternsWithFormula |>
  dplyr::filter(formula_name == "fixed amount formulation") |>
  dplyr::select(pattern_id, amount, amount_unit, numerator, numerator_unit, denominator, denominator_unit) |>
  flextable::flextable()
```

The formula in this case will be as follows:

\begin{equation}
\textrm{daily dose} = \frac{quantity \cdot amount}{days\: exposed}
\end{equation}

In this case unit will be determined by the `amount_unit` field.

4. **Concentration formulation**

This formula was applied for the following 30 patterns that cover 1,398,518 (47%) relationships:

```{r, echo = FALSE}
patternsWithFormula |>
  dplyr::filter(formula_name == "concentration formulation") |>
  dplyr::select(pattern_id, amount, amount_unit, numerator, numerator_unit, denominator, denominator_unit) |>
  flextable::flextable()
```

The formula in this case will be as follows:

\begin{equation}
\textrm{daily dose} = \frac{quantity \cdot numerator}{days\: exposed}
\end{equation}

In this case unit will be determined by the `numerator_unit` field.

For formulas (3) and (4) quantity is obtained from the `quantity` column of the *drug exposure* table and time exposed is obtained as the difference in days between `drug_exposure_start_date` and `drug_exposure_end_date` plus one.

The described formulas and patterns can be found in the exported `patternsWithFormula` data set:

```{r}
patternsWithFormula |>
  dplyr::glimpse()
```

The described formulas were validated into 5 different databases and the results were included in an article. Please refer to it for more details on dose calculations: [Calculating daily dose in the Observational Medical Outcomes Partnership Common Data Model](https://doi.org/10.1002/pds.5809).

## Finding out the pattern information using patternTable() function

The user could also find the patterns used in the `drug_strength` table. The output will also include a column of potentially valid and invalid combinations. The idea of a pattern to provide a platform to associate each drug in the `drug_strength` table with its constituent ingredients.

```{r, message= FALSE, warning=FALSE}
patternTable(cdm) |>
  dplyr::glimpse()
```

The output has three important columns, namely `number_concepts`, `number_ingredients` and `number_records`, which corresponds to count of distinct concepts in the patterns, count of distinct ingredients involved and overall count of records in the patterns respectively. The `pattern_id` column can be used to relate the patterns with the `patternsWithFormula` data set.

## Finding out the dose coverage using summariseDoseCoverage() function

This package also provides a functionality to check the coverage daily dose computation for chosen concept sets and ingredients. Let's take *acetaminophen* as an example.

```{r, message= FALSE, warning=FALSE}
summariseDoseCoverage(cdm = cdm, ingredientConceptId = 1125315) |>
  dplyr::glimpse()
```

The output will summarise the usage of *acetaminophen* in the database. For example, overall there are $78$ records of *acetaminophen* and for all of them daily dose can be calculated. By default the output will also include the mean, median, lower and upper quartiles and standard deviation of the daily dose of *acetaminophen* calculated as explained above. The results will also be stratified by unit, route and pattern (which we saw in `patternsWithFormula` data set).

Different routes are documented in the **CodelistGenerator** package. Route is defined at the concept (`drug_concept_id`) level, there exist an equivalence between each concept and a route. You can stratify a codelist using the function: `CodelistGenerator::stratifyByRouteCategory()`.

To better inspect the content of the output of `summariseDoseCoverage()` we can create a gt table like so:

```{r, message= FALSE, warning=FALSE}
coverageResult <- summariseDoseCoverage(cdm = cdm, ingredientConceptId = 1125315)
tableDoseCoverage(coverageResult)
```

The user also has the freedom to customize the gt table output. For example the following will suppress the `cdmName`:

```{r, message= FALSE, warning=FALSE}
tableDoseCoverage(coverageResult, cdmName = F)
```
