---
title: "Identify and summarise indications among a drug cohort"
output: 
  html_document:
    pandoc_args: [
      "--number-offset=1,0"
      ]
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Add indications to Drug Utilisation Cohort}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
if (Sys.getenv("EUNOMIA_DATA_FOLDER") == "") Sys.setenv("EUNOMIA_DATA_FOLDER" = tempdir())
if (!dir.exists(Sys.getenv("EUNOMIA_DATA_FOLDER"))) dir.create(Sys.getenv("EUNOMIA_DATA_FOLDER"))
if (!CDMConnector::eunomia_is_available()) CDMConnector::downloadEunomiaData()
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

In this vignette, we demonstrate the functionality provided by the DrugUtilisation package to help understand the indications of patients in a drug cohort.

The DrugUtilisation package is designed to work with data in the OMOP CDM format, so our first step is to create a reference to the data using the DBI and CDMConnector packages. 

```{r setup , message= FALSE, warning=FALSE}
library(DrugUtilisation)
library(DBI)
library(duckdb)
library(CDMConnector)
library(dplyr)
library(PatientProfiles)

con <- DBI::dbConnect(duckdb::duckdb(), eunomia_dir())

cdm <- CDMConnector::cdm_from_con(
  con = con,
  cdm_schema = "main",
  write_schema = "main"
)
```

## Create a drug utilisation cohort

We will use *acetaminophen* as our example drug. We'll start by creating a cohort of acetaminophen users. Here we'll include all acetaminophen records using a gap era of 7 days, but as we've seen in the previous vignette we could have also applied various other inclusion criteria.

```{r, eval=TRUE, message= FALSE, warning=FALSE}
cdm <- generateIngredientCohortSet(
  cdm = cdm,
  name = "acetaminophen_users",
  ingredient = "acetaminophen", 
  gapEra = 7
)
```
## Create a indication cohort

Next we will create a set of indication cohorts. In this case we will create cohorts for sinusitis  and bronchitis using `CDMConnector::generateConceptCohortSet()`. 

```{r, eval=TRUE, message= FALSE, warning=FALSE}
indications <-
  list(
    sinusitis = c(257012, 4294548, 40481087),
    bronchitis = c(260139, 258780)
  )

cdm <- generateConceptCohortSet(cdm, name = "indications_cohort", indications)
cdm
```


# Add indications with addIndication() function

Now that we have these two cohort tables, one with our drug cohort and another with our indications cohort, we can assess patient indications. For this we will specify a time window around the drug cohort start date for which we identify any intersection with the indication cohort. We can add this information as a new variable on our cohort table.

```{r, eval=TRUE, message= FALSE, warning=FALSE}
cdm[["acetaminophen_users"]] <- cdm[["acetaminophen_users"]] |>
  addIndication(
    indicationCohortName = "indications_cohort",
    indicationWindow =  list(c(-30,0)), 
    indexDate = "cohort_start_date"
  )
cdm[["acetaminophen_users"]] |> glimpse()
```

We can see that individuals are classified as having sinusistis (without bronchitis), bronchitis (without sinusitis), sinusitis and bronchitis, or no observed indication.
```{r, eval=TRUE, message= FALSE, warning=FALSE}
cdm[["acetaminophen_users"]] |> 
  group_by(indication_m30_to_0) |> 
  tally()
```

As well as the indication cohort table, we can also use the clinical tables in the OMOP CDM to identify other, unknown, indications. Here we consider anyone who is not in an indication cohort but has a record in the condition occurrence table to have an "unknown" indication. We can see that many of the people previously considered to have no indication are now considered as having an unknown indication as they have a condition occurrence record in the 30 days up to their drug initiation.

```{r, eval=TRUE, message= FALSE, warning=FALSE}
cdm[["acetaminophen_users"]] |>
  select(!"indication_m30_to_0") |> 
  addIndication(
    indicationCohortName = "indications_cohort",
    indicationWindow =  list(c(-30,0)),
    unknownIndicationTable = "condition_occurrence"
  ) |> 
  group_by(indication_m30_to_0) |> 
  tally()
```

We can add indications for multiple time windows. Unsurprisingly we find more potential indications for wider windows (although this will likely increase our risk of false positives).
```{r, eval=TRUE, message= FALSE, warning=FALSE}
cdm[["acetaminophen_users"]] <- cdm[["acetaminophen_users"]] |>
  select(!"indication_m30_to_0") |> 
  addIndication(
    indicationCohortName = "indications_cohort",
    indicationWindow =  list(c(0,0),
                             c(-30,0),
                             c(-365,0)),
    unknownIndicationTable = "condition_occurrence"
  )
cdm[["acetaminophen_users"]] |> 
  group_by(indication_0_to_0) |> 
  tally()
cdm[["acetaminophen_users"]] |> 
  group_by(indication_m30_to_0) |> 
  tally()
cdm[["acetaminophen_users"]] |> 
  group_by(indication_m365_to_0) |> 
  tally()
```

## Summarise indications with summariseIndication()

Instead of adding variables with indications like above, we could instead obtain a general summary of observed indications. `summariseIndication` has similar arguments to  `addIndication()`, but returns a summary result of the indication.

```{r, eval=TRUE, message= FALSE, warning=FALSE}
indication_summary <- cdm[["acetaminophen_users"]] |>
  select(!"indication_0_to_0") |> 
  select(!"indication_m30_to_0") |> 
  select(!"indication_m365_to_0") |> 
  summariseIndication(
    indicationCohortName = "indications_cohort",
    indicationWindow =  list(c(0,0),c(-30,0),c(-365,0)),
    unknownIndicationTable =  c("condition_occurrence")
  )
```

We can then easily create a plot or a table of the results
```{r, eval=TRUE, message= FALSE, warning=FALSE}
tableIndication(indication_summary)
```

As well as getting these overall results, we can also stratify the results by some variables of interest. For example, here we stratify our results by age groups and sex. 

```{r, eval=TRUE, message= FALSE, warning=FALSE}
indication_summary_stratified <- cdm[["acetaminophen_users"]] |>
  select(!"indication_0_to_0") |> 
  select(!"indication_m30_to_0") |> 
  select(!"indication_m365_to_0") |> 
  addDemographics(ageGroup = list(c(0, 19), c(20, 150))) |>
  summariseIndication(
    strata = list("age_group", "sex"),
    indicationCohortName = "indications_cohort",
    indicationWindow =  list(c(0,0),c(-30,0),c(-365,0)),
    unknownIndicationTable =  c("condition_occurrence")
  )
```

```{r, eval=TRUE, message= FALSE, warning=FALSE}
tableIndication(indication_summary_stratified)
```

