---
title: "Use DrugUtilisation to create cohort"
author: "Marti Catala, Mike Du, Yuchen Guo, Kim Lopez-Guell, Edward Burn, Xintong Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use DrugUtilisation to create cohort}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r , include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

In this vignette we will introduce different ways to create a cohort
using different functions of DrugUtilisation.

## Create mock data first

```{r setup, message = FALSE, warning = FALSE}
library(DrugUtilisation)
library(CodelistGenerator)
library(CDMConnector)
library(dplyr)

cdm <- mockDrugUtilisation(numberIndividual = 200)
```

# Get the concept code

We will first generate a cohort using a concept list. The concept list
can be obtained through different ways:

1.  Get from .json file

2.  Use concept code directly

3.  Get ingredient

4.  Get ATC code

## Concept list from a .json file

To create a cohort with a concept list from a .json file, we can use
`readConceptList()` function from this package,
or`codesFromConceptSet()` from CDMConnector package. Let's see an
example:

```{r}
conceptSet_json_1 <- readConceptList(here::here("inst/Concept"), cdm) 
conceptSet_json_2 <- codesFromConceptSet(here::here("inst/Concept"), cdm)

conceptSet_json_1
conceptSet_json_2
```

## Concept list listed directly

The concept list can also be created manually:

```{r}
#get concept using code directly
conceptSet_code <- list(asthma = 317009)
conceptSet_code
```

## Concept list of an ingredient

In case there is a certain ingredient of interest, the code can also be
obtained by `getDrugIngredientCodes()` from the package
CodelistGenerator.

```{r, message=FALSE, warning=FALSE}
#get concept by ingredient
conceptSet_ingredient <- getDrugIngredientCodes(cdm, name = "simvastatin")
conceptSet_ingredient
```

## Concept list of an ingredient

We can also obtain the ATC code by using `getATCCodes()` from
CodelistGenerator package.

```{r, message=FALSE, warning=FALSE}
#get concept from ATC codes
conceptSet_ATC <- getATCCodes(cdm, 
                              level = "ATC 1st", 
                              name = "ALIMENTARY TRACT AND METABOLISM")
conceptSet_ATC
```

# Create a cohort

Once we have the `conceptSet`, we can proceed to generate a cohort.
There are two functions in this package to do that:

1.  `generateConceptCohortSet()`: to generate a cohort for a certain
    list of concepts, which does not have to be a drug. This function is
    exported from CDMConnector package.

2.  `generateDrugUtilisationCohortSet()`: to generate a cohort of the
    drug use.

## generateConceptCohortSet

Let's try to use `generateConceptCohortSet()` to get the asthma cohort
using the `conceptSet_code`. We could also use `conceptSet_json_1` or
`conceptSet_json_2` to obtain the same result, as they are using the
same concept code.

```{r, message=FALSE, warning=FALSE}
cdm <- generateConceptCohortSet(cdm,
  conceptSet = conceptSet_code,
  name = "asthma_1",
  end = "observation_period_end_date",
  requiredObservation = c(10, 10),
  overwrite = TRUE
)
cdm$asthma_1
```

The count of the cohort can be assessed using `cohortCount()` from
CDMConnector

```{r, message=FALSE, warning=FALSE}
cohortCount(cdm$asthma_1)
```

Cohort attrition can be assessed using `cohortAttrition()` from
CDMConnector

```{r, message=FALSE, warning=FALSE}
cohortAttrition(cdm$asthma_1)
```

### end

Notice that we have specified an `end` input. This parameter sets how
the cohort end date is defined. Now it is changed to event end date to
demonstrate the difference from previous observation period end date.
See that now the cohort_end_date is different:

```{r, message=FALSE, warning=FALSE}
cdm <- generateConceptCohortSet(cdm,
  conceptSet = conceptSet_code,
  name = "asthma_2",
  end = "event_end_date",
  requiredObservation = c(10, 10),
  overwrite = TRUE
)
cdm$asthma_2
```

### requiredObservation

The `requiredObservation` parameter is a numeric vector of length 2,
that defines the number of days of required observation time prior to
index and post index for an event to be included in the cohort. Let's
check it now to see how reducing required observation affect the
**asthma_1** cohort.

```{r, message=FALSE, warning=FALSE}
cdm <- generateConceptCohortSet(cdm,
  conceptSet = conceptSet_code,
  name = "asthma_3",
  end = "observation_period_end_date",
  requiredObservation = c(1, 1),
  overwrite = TRUE
)
cdm$asthma_3

cohortCount(cdm$asthma_3)

cohortAttrition(cdm$asthma_3)
```

## generateDrugUtilisationCohortSet

Now let's try function `generateDrugUtilisationCohortSet()` to get the
drug cohort for ingredient *simvastatin*. This function has a lot more
options you can set. Let's try first to use the default settings:

```{r, message=FALSE, warning=FALSE}
cdm <- generateDrugUtilisationCohortSet(cdm,
  name = "dus_alleras",
  conceptSet = conceptSet_ingredient
)
cdm$dus_alleras

cohortCount(cdm$dus_alleras)

cohortAttrition(cdm$dus_alleras)
```

### imputeDuration and durationRange

The parameter `durationRange` specifies the range within which the
duration must fall, where duration will be calculated as:
<p style="text-align:center;">`duration = cohort_end_date - cohort_start_date + 1`</p>
The default value is `c(1, Inf)`. See that this parameter must be a
numeric vector of length two, with no NAs and with the first value equal
or bigger than the second one. It should be a numeric vector of length
two, with no NAs and the first value should be equal or smaller than the
second one. Duration values outside of `durationRange` will be imputed
using `imputeDuration`. `imputeDuration` can be set as: `none`(default),
`median`, `mean`, `mode` or an integer (count).

```{r, message=FALSE, warning=FALSE}
cdm <- generateDrugUtilisationCohortSet(cdm,
  name = "dus_step2_0_inf",
  conceptSet = conceptSet_ingredient,
  imputeDuration = "none",
  durationRange = c(0, Inf) # default as c(1, Inf)
)

cohortAttrition(cdm$dus_step2_0_inf)
```

### gapEra

The `gapEra` parameter defines the number of days between two continuous
drug exposures to be considered as a same era. Now let's change it from
0 to a larger number to see what happens.

```{r, message=FALSE, warning=FALSE}
cdm <- generateDrugUtilisationCohortSet(cdm,
  name = "dus_step3_alleras",
  conceptSet = conceptSet_ingredient,
  imputeDuration = "none",
  durationRange = c(0, Inf),
  gapEra = 30 # default as 0
)

cohortAttrition(cdm$dus_step3_alleras) %>% select(number_records, reason, excluded_records, excluded_subjects)
```

From the **dus_step3_alleras** cohort attrition, we can see that when
joining eras, it resulted in less records, compared to the
**dus_step2_0_inf** cohort, as exposures with less than 30 days gaps are
joined.

### priorUseWashout

The `priorUseWashout` parameter specifies the number of prior days
without exposure (often termed a 'washout') that are required. By
default, it is set to NULL, meaning no washout period is necessary. In
the example provided, we observe a reduction in the number of records in
**STEP 4** for cohort **dus_alleras_step4** due to the washout period
required, compared to the **dus_step3_alleras** cohort.

```{r, message=FALSE, warning=FALSE}
cdm <- generateDrugUtilisationCohortSet(cdm,
  name = "dus_alleras_step4",
  conceptSet = conceptSet_ingredient,
  imputeDuration = "none",
  durationRange = c(0, Inf),
  gapEra = 30,
  priorUseWashout = 30
)

cohortAttrition(cdm$dus_alleras_step4) %>% select(number_records, reason, excluded_records, excluded_subjects)

```

### priorObservation

The parameter `priorObservation` defines the minimum number of days of
prior observation necessary for drug eras to be taken into account. If
set to NULL, the drug eras are not required to fall within the
observation_period. In this example, there is a noticeable decrease in
the number of records for **dus_alleras_step5** cohort in **STEP 5**
when compared to the **dus_alleras_step4** cohort.

```{r, message=FALSE, warning=FALSE}
cdm <- generateDrugUtilisationCohortSet(cdm,
  name = "dus_alleras_step5",
  conceptSet = conceptSet_ingredient,
  imputeDuration = "none",
  durationRange = c(0, Inf),
  gapEra = 30,
  priorUseWashout = 30,
  priorObservation = 30
)

cohortAttrition(cdm$dus_alleras_step5) %>% select(number_records, reason, excluded_records, excluded_subjects)
```

### cohortDateRange

The `cohortDateRange` parameter defines the range for the
**cohort_start_date** and **cohort_end_date**. In the following example,
one can observe a reduction in **STEP 6** and **STEP 7** due to the
constraints imposed on the cohort start and end dates.

```{r, message=FALSE, warning=FALSE}
cdm <- generateDrugUtilisationCohortSet(cdm,
  name = "dus_alleras_step67",
  conceptSet = conceptSet_ingredient,
  imputeDuration = "none",
  durationRange = c(0, Inf),
  gapEra = 30,
  priorUseWashout = 30,
  priorObservation = 30,
  cohortDateRange = as.Date(c("2010-01-01", "2011-01-01")),
  limi = "All"
)

cohortAttrition(cdm$dus_alleras_step67) %>% select(number_records, reason, excluded_records, excluded_subjects)
```

### limit

The input `limit` allows `all` (default) and `first` options. If we set
it to `first`, we will only obtain the first record that fulfills all
the criteria. Observe how it impacts the attrition of the
**dus_step8_firstera cohort** in comparison to the
**dus_alleras_step67** cohort. The number of records decreased at **STEP
8** because of the `First` limit.

```{r, message=FALSE, warning=FALSE}
cdm <- generateDrugUtilisationCohortSet(cdm,
  name = "dus_step8_firstera",
  conceptSet = conceptSet_ingredient,
  imputeDuration = "none",
  durationRange = c(0, Inf),
  gapEra = 30,
  priorUseWashout = 30,
  priorObservation = 30,
  cohortDateRange = as.Date(c("2010-01-01", "2011-01-01")),
  limit = "First"
)

cohortAttrition(cdm$dus_step8_firstera) %>% select(number_records, reason, excluded_records, excluded_subjects)
```

If we just wanted to get the first-ever era, we can also use this
parameter. To achieve that, try the following setting:

```{r}
cdm <- generateDrugUtilisationCohortSet(cdm,
  name = "dus_step8_firstever",
  conceptSet = conceptSet_ingredient,
  imputeDuration = "none",
  durationRange = c(0, Inf),
  gapEra = 0,
  priorUseWashout = Inf,
  priorObservation = 0,
  cohortDateRange = as.Date(c(NA, NA)),
  limit = "First"
)
```

Constructing concept sets and generating various cohorts are the initial
steps in conducting a drug utilisation study. For further guidance on
using getting more information like characteristics from these cohorts,
please refer to the other vignettes.
