---
title: "Create mock data to test DrugUtilisation package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DrugUtilisation mock data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = FALSE}
library(DrugUtilisation)
```

# Introduction

In this vignette we will see how to use `mockDrugUtilisation()` function to create mock data. This function is predominantly used in this package's unit testing.

For example, one could use the default parameters to create a mock cdm reference like so:

```{r, message= FALSE, warning=FALSE}
cdm <- mockDrugUtilisation()
```

This will then populate several omop tables (for example, `person`, `concept` and `visit_occurrence`) and two cohorts in the cdm reference.

```{r, message= FALSE, warning=FALSE}
cdm$person |> 
  dplyr::glimpse()

cdm$person |>
  dplyr::tally()
```

```{r, message= FALSE, warning=FALSE}
cdm$concept |> 
  dplyr::glimpse()

cdm$concept |>
  dplyr::tally()
```

```{r, message= FALSE, warning=FALSE}
cdm$visit_occurrence |> 
  dplyr::glimpse()

cdm$visit_occurrence |>
  dplyr::tally()
```

```{r, message= FALSE, warning=FALSE}
cdm$cohort1 |> 
  dplyr::glimpse()

cdm$cohort1 |>
  dplyr::tally()
```

```{r, message= FALSE, warning=FALSE}
cdm$cohort2 |> 
  dplyr::glimpse()

cdm$cohort2 |>
  dplyr::tally()
```

## Setting seeds

The user can also set the seed to control the randomness within the data.

```{r, message= FALSE, warning=FALSE}
cdm <- mockDrugUtilisation(
  seed = 789
)
```

We now observe that `cohort1` has been changed as a result of this seed:

```{r, message= FALSE, warning=FALSE}
cdm$cohort1 |> 
  dplyr::glimpse()
```

The users can then create mock data in two ways, one is to set the `numberIndividual` parameter and the other is to cusutomise the tables.

## Create mock data using numberIndividual parameter

An example of use is as follows:

```{r, message= FALSE, warning=FALSE}
cdm <- mockDrugUtilisation(numberIndividual = 100)
```

This will ensure that each of `person`, `observation_period`, `cohort1` and `cohort2` will have 100 rows. 

```{r, message= FALSE, warning=FALSE}
cdm$person |>
  dplyr::glimpse()
```

```{r, message= FALSE, warning=FALSE}
cdm$person |>
  dplyr::tally()
```

As a consequence of this, the number of rows for other tables such as `visit_occurrence`, `condition_occurrence` and `drug_strength` will have more rows compared to the mock data produced using default settings.

```{r, message= FALSE, warning=FALSE}
cdm$visit_occurrence |>
  dplyr::glimpse()
```

```{r, message= FALSE, warning=FALSE}
cdm$visit_occurrence |>
  dplyr::tally()
```

## Creat mock data by customising tables

### Customise omop tables

As we saw previously, the omop tables are automatically populated in `mockDrugUtilisation()`. However, the user can customise these tables. For example, to customise `drug_exposure` table, one could do the following:

```{r, message= FALSE, warning=FALSE}
cdm <- mockDrugUtilisation(
    drug_exposure = dplyr::tibble(
      drug_exposure_id = 1:3,
      person_id = c(1, 1, 1),
      drug_concept_id = c(2, 3, 4),
      drug_exposure_start_date = as.Date(c(
        "2000-01-01", "2000-01-10", "2000-02-20"
      )),
      drug_exposure_end_date = as.Date(c(
        "2000-02-10", "2000-03-01", "2000-02-20"
      )),
      quantity = c(41, 52, 1),
      drug_type_concept_id = 0
    )
  )
```

```{r, message= FALSE, warning=FALSE}
cdm$drug_exposure |>
  dplyr::glimpse()
```

However, one needs to be vigilant that the customised omop table is implicitly dependent on other omop tables. For example, the `drug_exposure_start_date` of someone in the `drug_exposure` table should lie within that person's `observation_period_start_date` and `observation_period_end_date`.

One could also modify other omop tables including `person`, `concept`, `concept_ancestor`, `drug_strength`, `observation_period`, `condition_occurrence`, `observation`, and `concept_relationship` in a similar fashion.

### Customise cohorts

Similarly, cohort tables can also be customised.

```{r, message= FALSE, warning=FALSE}
cdm <- mockDrugUtilisation(
  observation_period = dplyr::tibble(
      observation_period_id = 1,
      person_id = 1:2,
      observation_period_start_date = as.Date("1900-01-01"),
      observation_period_end_date = as.Date("2100-01-01"),
      period_type_concept_id = 0
    ),
    cohort1 = dplyr::tibble(
      cohort_definition_id = 1,
      subject_id = c(1, 1, 2),
      cohort_start_date = as.Date(c("2000-01-01", "2001-01-01", "2000-01-01")),
      cohort_end_date = as.Date(c("2000-03-01", "2001-03-01", "2000-03-01"))
    )
  )
```

```{r, message= FALSE, warning=FALSE}
cdm$cohort1 |> 
  dplyr::glimpse()
```
