---
title: "Create mock data to test DrugUtilisation package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DrugUtilisation mock data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = FALSE}
library(DrugUtilisation)
```

# Introduction

In this vignette we will see how to use `mockDrugUtilisation()` function to create mock data. This function is predominantly used in this package's unit testing. But it can also be useful to create some mock data to learn some of the functionalities of the package.

For example, one could use the default parameters to create a mock cdm reference like so:

```{r, message= FALSE, warning=FALSE}
cdm <- mockDrugUtilisation()
```

This will then populate several omop tables (for example, `person`, `concept` and `visit_occurrence`) and two cohorts in the cdm reference.

```{r, message= FALSE, warning=FALSE}
cdm$person |> 
  dplyr::glimpse()

cdm$person |>
  dplyr::tally()
```

```{r, message= FALSE, warning=FALSE}
cdm$concept |> 
  dplyr::glimpse()

cdm$concept |>
  dplyr::tally()
```

```{r, message= FALSE, warning=FALSE}
cdm$visit_occurrence |> 
  dplyr::glimpse()

cdm$visit_occurrence |>
  dplyr::tally()
```

```{r, message= FALSE, warning=FALSE}
cdm$cohort1 |> 
  dplyr::glimpse()

cdm$cohort1 |>
  dplyr::tally()
```

```{r, message= FALSE, warning=FALSE}
cdm$cohort2 |> 
  dplyr::glimpse()

cdm$cohort2 |>
  dplyr::tally()
```

## Setting seeds

The user can also set the seed to control the randomness within the data. This would allow us to regenerate the same mock data again.

```{r, message= FALSE, warning=FALSE}
cdm <- mockDrugUtilisation(
  seed = 789
)
```

Note that to consecutive calls of `mockDrugUtilisation()` do not generate the same mock data, if the seed parameter is not set to a specific number. For example we can now observe that `cohort1` is different as a result of this seed:

```{r, message= FALSE, warning=FALSE}
cdm$cohort1 |> 
  dplyr::glimpse()
```

The users can then create mock data in two ways, one is to set the `numberIndividual` parameter and the other is to cusutomise the tables.

## Create mock data using numberIndividual parameter

An example of use is as follows:

```{r, message= FALSE, warning=FALSE}
cdm <- mockDrugUtilisation(numberIndividual = 100)
```

This will ensure that each of `person`, `observation_period`, `cohort1` and `cohort2` will have 100 rows. 

```{r, message= FALSE, warning=FALSE}
cdm$person |>
  dplyr::glimpse()
```

```{r, message= FALSE, warning=FALSE}
cdm$person |>
  dplyr::tally()
```

As a consequence of this, the number of rows for other tables such as `visit_occurrence`, `condition_occurrence` and `drug_exposure` will have more rows compared to the mock data produced using default settings.

```{r, message= FALSE, warning=FALSE}
cdm$visit_occurrence |>
  dplyr::glimpse()
```

```{r, message= FALSE, warning=FALSE}
cdm$visit_occurrence |>
  dplyr::tally()
```

## Creat mock data by customising tables

### Customise omop tables

As we saw previously, the omop tables are automatically populated in `mockDrugUtilisation()`. However, the user can customise these tables. For example, to customise `drug_exposure` table, one could do the following:

```{r, message= FALSE, warning=FALSE}
cdm <- mockDrugUtilisation(
    drug_exposure = dplyr::tibble(
      drug_exposure_id = 1:3,
      person_id = c(1, 1, 1),
      drug_concept_id = c(2, 3, 4),
      drug_exposure_start_date = as.Date(c(
        "2000-01-01", "2000-01-10", "2000-02-20"
      )),
      drug_exposure_end_date = as.Date(c(
        "2000-02-10", "2000-03-01", "2000-02-20"
      )),
      quantity = c(41, 52, 1),
      drug_type_concept_id = 0
    )
  )
```

```{r, message= FALSE, warning=FALSE}
cdm$drug_exposure |>
  dplyr::glimpse()
```

However, one needs to be vigilant that the customised omop table is implicitly dependent on other omop tables. For example, the `drug_exposure_start_date` of someone in the `drug_exposure` table should lie within that person's `observation_period_start_date` and `observation_period_end_date`. And that individual should be present in person table with a birth date before the `drug_exposure_start_date`.

One could also modify other omop tables including `person`, `concept`, `concept_ancestor`, `drug_strength`, `observation_period`, `condition_occurrence`, `observation`, and `concept_relationship` in a similar fashion.

### Customise cohorts

Similarly, cohort tables can also be customised.

```{r, message= FALSE, warning=FALSE}
cdm <- mockDrugUtilisation(
  cohort1 = dplyr::tibble(
    cohort_definition_id = 1,
    subject_id = c(1, 1, 2),
    cohort_start_date = as.Date(c("2000-01-01", "2001-01-01", "2000-01-01")),
    cohort_end_date = as.Date(c("2000-03-01", "2001-03-01", "2000-03-01"))
  )
)
```

```{r, message= FALSE, warning=FALSE}
cdm$cohort1 |> 
  dplyr::glimpse()
```

Automatically the person table will be populated with individuals 1 and 2:
```{r}
cdm$person |>
  dplyr::glimpse()
```

And each one of this subjects will have an observation period that starts before the first cohort observation and ends after the last cohort observation.
```{r}
cdm$observation_period |>
  dplyr::glimpse()
```

## Insert you cdm object to a database

By default the cdm reference object provided is in duckdb:
```{r}
omopgenerics::cdmSourceType(cdm)
```

You can insert this to another database management system, in this case if we would have an instance of sql server we could insert it like:
```{r, eval = FALSE}
con <- DBI::dbConnect(
  odbc::odbc(),
  Driver   = "ODBC Driver 18 for SQL Server",
  Server   = Sys.getenv("SQL_SERVER_SERVER"),
  Database = Sys.getenv("SQL_SERVER_CDM_DATABASE"),
  UID      = Sys.getenv("SQL_SERVER_USER"),
  PWD      = Sys.getenv("SQL_SERVER_PASSWORD"),
  TrustServerCertificate = "yes",
  Port     = Sys.getenv("SQL_SERVER_PORT")
)
cdm <- mockDrugUtilisation(con = con, writeSchema = c(catalog = "omop", schema = "results", prefix = "mock_dus_"))
```

This would write the mock tables in the `results` schema of `omop` catalog, and all table names would start with "mock_dus_".
```{r, echo = FALSE}
con <- DBI::dbConnect(duckdb::duckdb())
cdm <- mockDrugUtilisation(
  con = con, writeSchema = c(schema = "main", prefix = "mock_dus_"))
attributes(attr(cdm, "cdm_source"))$source_type <- "sql server"
```
```{r}
omopgenerics::cdmSourceType(cdm)
DBI::dbListTables(con)
```
