---
title: "Assessing time to treatment discontinuation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assessing time to treatment discontinuation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
if (Sys.getenv("EUNOMIA_DATA_FOLDER") == "") Sys.setenv("EUNOMIA_DATA_FOLDER" = tempdir())
if (!dir.exists(Sys.getenv("EUNOMIA_DATA_FOLDER"))) dir.create(Sys.getenv("EUNOMIA_DATA_FOLDER"))
if (!CDMConnector::eunomia_is_available()) CDMConnector::downloadEunomiaData()
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Time-to-discontinuation is commonly of interest when summarising adherence to  medications. 

# Time-to-discontinuation of amoxicillin
For example, let's say we would like to study time-to-discontinuation among new users of amoxicillin in their first 90-days of use. For this we can first create our amoxicillin study cohort, restricting to first occurrence in history. We can then estimate Kaplan-Meier survival curves we can use estimateSingleEventSurvival from the CohortSurvival package, with the follow-up beginning from an individual's cohort start date and discontinuation occurring at their cohort end date. 

Here we'll use the synthetic Eunomia dataset to show how this could be done.

```{r}
library(CDMConnector)
library(PatientProfiles)
library(DrugUtilisation)
library(CohortSurvival)
library(dplyr)

db <- DBI::dbConnect(duckdb::duckdb(), 
                     dbdir = eunomia_dir())
cdm <- cdm_from_con(
  con = db,
  cdm_schema = "main", 
  write_schema = "main", 
  cdm_name = "Eunomia"
)

cdm <- generateIngredientCohortSet(cdm = cdm,
                                   name = "amoxicillin",
                                   ingredient = "amoxicillin", 
                                   gapEra = 7) 
cdm$amoxicillin <- cdm$amoxicillin |> 
  requireIsFirstDrugEntry()

discontinuation_summary <- estimateSingleEventSurvival(cdm, 
                                    targetCohortTable = "amoxicillin", 
                                    outcomeCohortTable = "amoxicillin", 
                                    outcomeDateVariable = "cohort_end_date", 
                                    followUpDays = 90, 
                                    eventGap = 30)
```

We can plot our study result like so:
```{r, out.width="75%"}
plotSurvival(discontinuation_summary)
```

Or we can similarly create a table summarising the result
```{r}
tableSurvival(discontinuation_summary)
```

We can also easily stratify our results. Here we add patient demographics to our cohort table using the PatientProfiles packages and then stratify results by age group and sex.
```{r}
cdm$amoxicillin <- cdm$amoxicillin |> 
  addDemographics(ageGroup = list(c(0, 40),
                                  c(41, Inf))) |> 
  compute(temporary = FALSE, name = "amoxicillin")

discontinuation_summary <- estimateSingleEventSurvival(cdm, 
                                                       strata = list(c("age_group"),
                                                                     c("sex")),
                                    targetCohortTable = "amoxicillin", 
                                    outcomeCohortTable = "amoxicillin", 
                                    outcomeDateVariable = "cohort_end_date", 
                                    followUpDays = 90, 
                                    eventGap = 30)
```

Again we could present our results in a plot or a table.
```{r, out.width="75%"}
plotSurvival(discontinuation_summary, 
             facet = "strata_level")
```

```{r}
tableSurvival(discontinuation_summary)
```
